---
title: "Data Analysis"
author: "Parthasarathy Krishnamurthy"
output: html_document
---

##Overview

The objective of this exercise is to use confidence intervals to compare data. For this we will be using a dataset provided by R programming language. 


##Data load

The ToothGrowth dataset is loaded.

```{r}
library(datasets)
data(ToothGrowth)
summary(ToothGrowth)
```

The above result is basic summary of the ToothGrowth dataset.

Let's look at some sample data.

```{r}
head(ToothGrowth)
```

To know thw unique values of supplementary type and dosage.

```{r}
unique(ToothGrowth$supp)
unique(ToothGrowth$dose)
```

We plot some of the important features to perform data analysis.

```{r}
library(lattice)
xyplot(len~dose|supp, ToothGrowth, main="Tooth length by dose for each supplement type", ylab="Length", xlab="Dose")
```

The basic analysis of data tells us that it represents the length of the teeth at various dosage levels of vitamins for each of the 2 delivery methods (VC and OJ).

##Tooth growth by supplementary type and dosage.

We will perform this analysis using the t-test.

First lets obtain the difference between groups of data with different supplementary types. Lets assume unequal variances between the 2 groups.

```{r}
t.test(len ~ supp, data = ToothGrowth)
```

The p value is 0.06 which is greater than 0.05 and confidence levels moves from negative to positive. So the confidence level has a value of zero, so based on this test the supplement types have no impact on tooth growth.

Next lets compare tooth growth by dosage.

```{r}
t.test(len ~ dose, data=subset(ToothGrowth, dose %in% c(0.5, 1.0)))
t.test(len ~ dose, data=subset(ToothGrowth, dose %in% c(0.5, 2.0)))
t.test(len ~ dose, data=subset(ToothGrowth, dose %in% c(1.0, 2.0)))
```

For all the 3 tests above, the p value is less than 0.05. Also none of the confidence levels move from negative to positive which means they do not have a zero value. This suggests that on increasing the dosage, the tooth length also increases. 


##Conclusion

The conclusions are based on the following assumptions:

1. The sample is representative of the population.
2. The supplement type is independent of dosage.

In conclusion, the supplement type has no effect on the tooth length (or growth). The mean tooth length increases on raising the dosage.